The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-06 handnotes.tex
edited variables. added some notes to random walk section.

  Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. Thus we say that some function $f(t)$ is a phenotype where, 
  \begin{align}
    f(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.

  Essentially the phenotype $f(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.

@@

  Although $f(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
  \begin{align}
    \Sigma = \left\{ \begin{array}{ll} \dot{x}(t) &= A x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right.
  \end{align}
  Variables have the same identities as described above and $x(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $y(t)$, are ``observed'' by selection.

  \begin{example}[Oscillating Gene Network: Circadian Rhythm]
    \begin{align*}
      A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
    \end{align*}
    \begin{align*}
      \Sigma = \left \{ \begin{array}{ll} \dot{x}(t) &= \begin{bmatrix} 
        0 & 1 \\ 
       -1 & 0 
        \end{bmatrix} x (t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
          y(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} x(t) \end{array} \right.
     \end{align*}

@@

    \begin{figure}[H]
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$x_{1}$};
                \node[dashed] (B) at (3,0) {$x_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

@@

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{x}(t) &= A(\tau) x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right .
  \end{align}
  \begin{example}[External Equivalence does not imply internal equivalence]
    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
  \end{figure}
  \end{example}
  Evolution thus proceeds as a random walk in phenotypically invariant network space.

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}
  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$.

  \begin{example}[All Phenotypically Equivalent Oscillators]
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix} \\
        B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
          h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

    Presently we ask two questions, (1) holding $f(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?

@@

      Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
      \begin{align}
        h(t) &= \widehat{h}(t) \iff \\ 
        CA^{k}B &= \widehat{C} \widehat{A}^{k} \widehat{C} \qquad \text{for } k = 0, 1, \dots
      \end{align}
      Any two systems with equivalent impulse responses will have equivalent phenotypes.
      \end{example}

@@

      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
      \begin{align}
        \Phi\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
      \end{align}
      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.

>>>>

  Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. Thus we say that some function $\phi(t)$ is a phenotype where, 
  \begin{align}
    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.

  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.

@@

  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
  \begin{align}
    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}
  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).

  \begin{example}[Oscillating Gene Network: Circadian Rhythm]
    \begin{align*}
      A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
    \end{align*}
    \begin{align*}
      \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
        0 & 1 \\ 
       -1 & 0 
        \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
          \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
     \end{align*}

@@

    \begin{figure}[H]
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

@@

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  \begin{example}[External Equivalence does not imply internal equivalence]
    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
  \end{figure}
  \end{example}
  Evolution thus proceeds as a random walk in phenotypically invariant network space.

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}
  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$. 
\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
  \begin{align}
    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  \end{align}
  \begin{example}[All Phenotypically Equivalent Oscillators]
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix} \\
        B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
          h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

    Presently we ask two questions, (1) holding $\phi(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?

@@

      Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
      \begin{align}
        h(t) &= \widehat{h}(t) \iff \\ 
        CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
      \end{align}
      Any two systems with equivalent impulse responses will have equivalent phenotypes.
      \end{example}

@@

      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
      \begin{align}
        \mathcal{F}\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
      \end{align}
      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.
